Follow-up Samples

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline methods and figure legends

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyperparameter involved taking the samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created.

***Initial Follow Up Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The training set utilized the `r length(rownames(good_metaf))`-person cohort that was previously used for testing of the lesion model.  The creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or initial follow up models.

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al. [@delong_comparing_1988] executed by the pROC (v`r paste(packageVersion("pROC"))`) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

***Analysis Overview***
Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.  All OTUs used in the lesion model were then also analyzed using a paired wilcoxson test.  We then investigated the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.  Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.  We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

@@

**Figure 1: Changes and differences between the adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the intial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial/Follow up models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set.  A) Lesion model.  B) Initial/Follow up model

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Reduced Variable Models**  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.

@@

**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Full Variable Model** A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.

>>>>

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***FIT and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance.  Data columns that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyperparameter involved taking the samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created.

***Initial Follow Up Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort follow up data set was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.

***Selection of Important OTUs***
In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or initial follow up models.

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in the lesion model were then analyzed using a paired wilcoxson test.  We then investigated the relative abundance of specific previously CRC associated bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial follow up model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

@@

**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the intial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classfied as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial and Follow up Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set.  A) Lesion model.  B) Initial/Follow up model

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Follow Up Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.    D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up Full Variable Models** A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.    D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline methods and figure legends

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the model used for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}21{]}. Kept apart from this training set were a total of 67
individuals that not only had a sample as described previoulsy but also
a follow up sample between 188 - 546 days after surgery and treatment
had been completed. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) {[}22{]}. 16S rRNA
gene sequencing was completed as previously described by Kozich, et al.
{[}23{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
system (Eppendorf). The V4 variable region was amplified and the
resulting product was split between three sequencing runs with control,
adenoma, and carcinoma evenly represented on each run. Each group was
randomly assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}23{]}. The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) for the 67 individuals with follow up samples.
The model included data on FIT and the bacterial microbiome. Non-binary
data was checked for near zero variance and auto correlation. Data
columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v6.0.73). Optimization of the mtry
hyperparameter involved taking the samples and making 100 80/20
(train/test) splits in the data where control and lesion were equally
represented in the 80 and 20 split, respectively. This 80\% portion was
then split again into an 80/20 split, and run through 20 repeated
10-fold cross validations to optimzie the model's AUC (Area Under the
Curve of the Receiver Operator Characteristic). This resulting model was
then tested on the 20\% of the data that was originally held out from
this overall process. Once the ideal mtry was found the entire 490
sample set was used to create the final Random Forest model on which
testing on the 67-person cohort was completed. The default cutoff of 0.5
was used as the threshold to classify individuals as positive or
negative for lesion. The hyperparameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The training set utilized the 67-person cohort that was
previously used for testing of the lesion model. The creation of this
model and optimization of the mtry hyperparameter was completed using
the same procedure that was used to create the lesion model.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion or initial follow up models.

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}25{]} in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. {[}26{]} executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied {[}27{]} and a corrected P-value of less than
0.05 was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
follow ups for either adenoma or carcinoma were investigated. Next,
initial and follow up samples were analyzed for differences in alpha and
beta diversity. All OTUs used in the lesion model were then also
analyzed using a paired wilcoxson test. We then investigated the
relative abundance of specific bacteria, that have been previously
associated with CRC, we selected OTUs that taxonomically classified to
Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
test if there were any differences based on whether the individual had
an adenoma or carcinoma. The lesion model was then tested for accuracy
in prediction and whether it reduced the positive probability of lesion
after surgery. The most important OTUs for this were used to build an
updated model and this reduced feature model was assessed for it's
similarity to the original model. We then used the initial follow up
model to assess whether this model could classify samples better then
the lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial follow up model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. Finally, a list of common OTUs
were found for the two different models used.

@@

\textbf{Figure 1: Changes and differences between the adenoma or
carcinoma group.} A) No significant difference was found between the
adenoma and carcinoma group for thetayc (P-value = 0.00198). B) A
significant difference was found between the adenoma and carcinoma group
for FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up
samples for the Adenoma group. D) NMDS of the initial and follow up
samples for the Carcinoma group. For C) and D) the teal represents
initial samples and the pink represents follow up samples.

\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
Follow up Samples.} A) Carcinoma initial and follow up samples. There
was a significant difference in initial and follow up sample for the
OTUs classfied as Peptostreptococcus stomatis (P-value = 0.0183) and
Porphyromonas asaccharolytica (P-value = 0.0154). B) Adenoma initial and
follow up samples. There were no significant differences between initial
and follow up.

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
for lesion and Initial/Follow up models.} The shaded areas represents
the range of values of a 100 different 80/20 splits of the test set data
using either all variables (grey) or reduced variable (red) models. The
blue line represents the reduced variable model using 100\% of the data
set. A) Lesion model. B) Initial/Follow up model

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial and Follow Up for Reduced Variable
Models} A) Positive probability adjustment of those with carcinoma from
intial to follow up sample B) Positive probability adjustment of those
with adenoma as well as those with SRN and the probability adjustment
from initial to follow up sample. The dotted line represents the
threshold used to make the decision of whether a sample was lesion
positive or not.

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial and Follow Up for Full Variable Model} A)
Positive probability adjustment of those with carcinoma from intial to
follow up sample B) Positive probability adjustment of those with
adenoma as well as those with SRN and the probability adjustment from
initial to follow up sample. The dotted line represents the threshold
used to make the decision of whether a sample was lesion positive or
not.

@@

\hypertarget{ref-delong_comparing_1988}{}
26. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
two or more correlated receiver operating characteristic curves: A
nonparametric approach. Biometrics. 1988;44:837--45.

\hypertarget{ref-benjamini_controlling_1995}{}
27. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.

>>>>

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}21{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described previoulsy but also a follow up
sample between 188 - 546 days after surgery and treatment had been
completed. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}23{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with control, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}23{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance. Data columns that had near zero variance were removed.
This pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyperparameter involved taking the samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
through 20 repeated 10-fold cross validations to optimzie the model's
AUC (Area Under the Curve of the Receiver Operator Characteristic). This
resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. Once the ideal mtry was
found the entire 490 sample set was used to create the final Random
Forest model on which testing on the 67-person cohort was completed. The
default cutoff of 0.5 was used as the threshold to classify individuals
as positive or negative for lesion. The hyperparameter, mtry, defines
the number of variables to investigate at each split before a new
division of the data is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The main difference was that only the 67-person cohort
follow up data set was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyperparameter was completed using the same
procedure that was used to create the lesion model.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most important to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion or initial follow up models.

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}25{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
wilcoxson ranked sum test. Where multiple comparison testing was needed
a Benjamini-Hochberg (BH) correction was applied {[}26{]} and a
corrected P-value of less than 0.05 was considered significant. Unless
otherwise stated the P-values reported are those of that were BH
corrected.

\textbf{\emph{Analysis Overview}} Initial and follow up samples were
analyzed for differences in alpha and beta diversity. Next, differences
in FIT between initial and follow ups for either adenoma or carcinoma
were investigated. From here, all OTUs that were used in the lesion
model were then analyzed using a paired wilcoxson test. We then
investigated the relative abundance of specific previously CRC
associated bacteria, specifically, OTUs that taxonomically classified to
\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}. We wanted to test if there were any differences based on
whether the individual had an adenoma or carcinoma. From here the lesion
model was then tested for accuracy in prediction and whether it reduced
the positive probability of lesion after surgery. The most important
OTUs for this were used to build a reduced model and it was assessed for
similarity to the original model. We then used the initial follow up
model to assess whether it could classify samples better then the lesion
model. The most important OTUs were then identified from this model and
used to create a reduced feature initial follow up model. This reduced
feature model, as was done with the lesion model, was compared to the
full model for loss of accuracy. Finally, a list of common OTUs were
found for the two different models used.

@@

\textbf{Figure 1: General Differences between the Adenoma or Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.00198). B) A significant
difference was found between the adenoma and carcinoma group for change
in FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up samples
for the Adenoma group. D) NMDS of the initial and follow up samples for
the Carcinoma group. For C) and D) the teal represents initial samples
and the pink represents follow up samples.

\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
Follow up Samples.} A) Carcinoma initial and follow up samples. There
was a significant difference in initial and follow up sample for the
OTUs classfied as \emph{Peptostreptococcus stomatis} (P-value = 0.0183)
and \emph{Porphyromonas asaccharolytica} (P-value = 0.0154). B) Adenoma
initial and follow up samples. There were no significant differences
between initial and follow up.

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
for lesion and Initial and Follow up Models.} The shaded areas
represents the range of values of a 100 different 80/20 splits of the
test set data using either all variables (grey) or reduced variable
(red) models. The blue line represents the reduced variable model using
100\% of the data set. A) Lesion model. B) Initial/Follow up model

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial Follow Up Reduced Variable Models} A)
Lesion positive probability adjustment of those with carcinoma from
initial to follow up sample B) Initial follow up positive probability
adjustment of those with carcinoma from initial to follow up sample C)
Lesion positive probability adjustment of those with adenoma as well as
those with SRN and the probability adjustment from initial to follow up
sample. D) Initial follow up positive probability adjustment of those
with adenoma as well as those with SRN and the probability adjustment
from initial to follow up sample. The dotted line represents the
threshold used to make the decision of whether a sample was positive or
not.

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial and Follow Up Full Variable Models} A)
Lesion positive probability adjustment of those with carcinoma from
initial to follow up sample B) Initial follow up positive probability
adjustment of those with carcinoma from initial to follow up sample C)
Lesion positive probability adjustment of those with adenoma as well as
those with SRN and the probability adjustment from initial to follow up
sample. D) Initial follow up positive probability adjustment of those
with adenoma as well as those with SRN and the probability adjustment
from initial to follow up sample. The dotted line represents the
threshold used to make the decision of whether a sample was positive or
not.

@@

\hypertarget{ref-benjamini_controlling_1995}{}
26. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
